---
title: "Introduction to sftime"
author: "Benedikt Gräler, Edzer Pebesma"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to sftime}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The package sftime extends the package sf with time support to store and handle spatiotemporal data. Following the concept of simple features in sf, sftime introduces a dedicated time column that is sued to provide the temporal information. The time column can consists of any collection of a class that allows to be sorted - reflecting the native order of time. Besides well-known time classes such as `Date` or `POSIXct`, it also allows for custom class definitions, that provide the necessary dedicated functions to make the sorting procedures work (we will see a example below).

This vignette is meant to briefly explain and illustrate the ideas and decisions behind the implementations of sftime.

## The time column

The time column is the special column of the underlying `data.frame` which defines the timestamps and the temporal ordering. Common time representations in R (e.g. "POSIXct", "POSIXlt", "Date", "yearmon", "yearqtr", "factor") are allowed, as well as the optional user defined type. Lets look at a simple example where we define a time column based on "POSIXct"

```{r}
library(sftime)
tc <- st_tc(as.POSIXct("2020-09-01 08:00:00")-0:3*3600*24)

tc
```

The ordering is not altered upon construction (as in some other representations). If a different order is required, the "order" and "sort" functions can be applied to the time column:

```{r}
tc[1]

order(tc)

sort(tc)
```

In some applications it might be useful to have more complex temporal information such as intervals of different length. The following example is also meant as template for other user defined classes which could be used to build the time column of the sftime class.

At first, we will need a few helper functions:

```{r}
# utility functions
as.character.interval <- function(x) {
  paste0("[", x[1], ", ", x[2], "]")
}

print.interval<- function(x) {
  cat("Interval:", as.character(x))
}

'[.intervals' <- function(x, i) {
  sx <- unclass(x)[i]
  class(sx) <- "intervals"
  sx
}
```

Now, we can define the different intervals used to represent our temporal information:

```{r}
# time interval definition
i1 <- c(5.3,12)
class(i1) <- "interval"
i2 <- c(3.1,6)
class(i2) <- "interval"
i3 <- c(1.4,6.9)
class(i3) <- "interval"
i4 <- c(1,21)
class(i4) <- "interval"

intrvls <- list(i1, i2, i3, i4)
class(intrvls) <- c("intervals", class(intrvls))
# provide dedicated generic to xtfrm for class intervals
xtfrm.intervals <- function(is) sapply(is, xtfrm)
```

The advantage is to be able to define different sorting approaches:

```{r}
# - sort by centre
xtfrm.interval <- function(i) mean(i)

tc <- st_tc(intrvls)
tc
tc[1]
order(tc)
sort(tc)[1]
```

```{r}
# - sort by end
xtfrm.interval <- function(i) i[2]
tc <- st_tc(intrvls)
tc
tc[1]
order(tc)
sort(tc)[1]
```

```{r}
# - sort by start
xtfrm.interval <- function(i) i[1]

tc <- st_tc(intrvls)
tc
tc[1]
order(tc)
sort(tc)[1]
```

Based on the sorting procedure (begin, centre or end of the interval), the smallest element (each last line) and the order of the time column changes.

## The sftime class

The class `sftime` is tightly coupled with the `sf` class and only introduces the additional time column. In this long-table style, it is ideal for irregular time series, but can serve any regular layout as well. However, the package `stars` is developed as a powerful alternative for regular spatiotemporal data sets (e.g. time series of remote sensing imagery, regular measurements of entire measurement network).

```{r}
g = st_sfc(st_point(1:2), st_point(c(1,3)), st_point(2:3), st_point(c(2,1)))

# provide as.data.frame method. This is needed for sf::st_sf to not cause an error with lists when calling as.data.frame
as.data.frame.intervals <- function(x, ...) {
  nm <- deparse1(substitute(x))
  if (!"nm" %in% names(list(...))) 
    as.data.frame.vector(x, ..., nm = nm)
  else as.data.frame.vector(x, ...)
}

sft <- st_sftime(a = 1:4, g, time = sort(tc))

sft <- st_sftime(g, time = st_tc(Sys.time()-0:3*3600*24))
```

This is the very beginning there is more to come.

## coercion

## sub-setting

## plotting
One of the first thing to do after a data set has been created, is to inspect it by plotting it. 
Plotting an 3+ dimensional onject is always a projection of selected dimensions. 
The following plotting abilities are provided in `sftime` through the mode argument (just as in `spacetime::stplot`). Depending on the nature of the data/underlying processes, some options might not lead to useful visulaisations:

  - `"xy"`: maps for time ranges are plotted, ranges can be defined or will be equally distributed. Geometries might be overlaid several times when multiple times fall in the same selected range.
  - `"xt"`: a space-time plot is constructed, this might lead to a straight line along the main diagonal in case not two geometries are recorded for the same time stamp nor are geometries recorded more than once. 
  - `"ts"`: multiple-locations time series are plotted in a single plot, or in a separate panel for each attribute
  - `"tp"`: single- or multi-attribute time series are plotted in multiple panels, one panel per location.
  
Examples for `mode="xy"`, the default:

```{r, fig.width=7}
coords <- matrix(runif(100), ncol = 2)
g = st_sfc(lapply(1:50, function(i) st_point(coords[i,]) ))
sf <- st_sf(a=1:50, g)

sft <- st_sftime(cbind(sf, time = st_tc(as.POSIXct("2020-09-01 00:00:00")+0:49*3600*6)))

plot(sft, key.pos = 4)

plot(sft, number=10, max.plot=10)

```


## ...
